Capital Budget Funding in Toronto

### 2024-09-26 paper/paper.qmd
Included summary results table and updated pdf

---
title: "Capital Budget Funding in Toronto"
subtitle: "An Analysis of the Funding Disparties Across Toronto Wards (2022 - 2031)"
author: 
  - Maria Mangru 
thanks: "Code and data are available at: https://github.com/MariaMangru/Capital-Budget-Allocations-and-Median-Household-Income."
date: today
date-format: long
abstract: "This research examines the allocation of capital funding across Toronto's 25 wards from 2022 to 2031, examining its relationship with socioeconomic indicators. The results reveal significant disparities in per capita funding, with higher-income wards and those with higher educational attainment receiving disproportionately more investment. Conversely, wards with higher proportions of low-income households and visible minority populations receive less funding per capita."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Despite the city's commitment to equity, recent studies have raised concerns about the fairness of budget distributions. Research by Walks et al. (2016) \[REFERENCE THIS\] highlights growing income inequality in Toronto's neighborhoods, noting a trend toward the spatial concentration of wealth and poverty. Similarly, Hulchanski (2010) \[REFERENCE THIS\] discusses the "Three Cities" within Toronto, where socioeconomic divisions are becoming more pronounced.

@@

    \newline

    This dataset provides comprehensive socioeconomic information for each ward based on the 2021 census. It was filtered to remain the variables which were most useful for this research. These are: Total Population, Median Household Income, Prevalence of Low Income Households, Educational Attainment, Immigration Status, Visible Minority Population. \[may need to include others, depends on what I decide to go with in the end.\]

>>>>

---
title: "Capital Budget Funding in Toronto"
subtitle: "An Analysis of the Funding Disparties Across Toronto Wards (2022 - 2031)"
author: Maria Mangru 
thanks: "Code and data are available at: https://github.com/MariaMangru/Capital-Budget-Allocations-and-Median-Household-Income."
date: today
date-format: long
abstract: "This research examines the allocation of capital funding across Toronto's 25 wards from 2022 to 2031, examining its relationship with socioeconomic indicators. The results reveal significant disparities in per capita funding, with higher-income wards and those with higher educational attainment receiving disproportionately more investment. Conversely, wards with higher proportions of low-income households and visible minority populations receive less funding per capita."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Despite the city's commitment to equity, recent studies have raised concerns about the fairness of budget distributions. Research by Walks et al. (2016) **\[REFERENCE THIS\]** highlights growing income inequality in Toronto's neighborhoods, noting a trend toward the spatial concentration of wealth and poverty. Similarly, PAPER HERE **\[REFERENCE THIS\]** WHAT IT TALKED ABOUT.

@@

This dataset provides comprehensive socioeconomic information for each ward based on the 2021 census. It was filtered to remain the variables which were most useful for this research. These are: Total Population, Median Household Income, Prevalence of Low Income Households, Educational Attainment, Immigration Status, Visible Minority Population. **\[UPDATE THIS\]**

```{r setting up workspace, echo=FALSE, message=FALSE, warning=FALSE}
#### Workspace setup ####
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(kableExtra)
library(scales)
library(ggplot2)
library(stringr)
library(janitor)

### Data Handling ###

# Read in the cleaned budget data
budget_data <- read_csv("../data/analysis_data/clean_budget_data.csv")

# Read in the cleaned demographic data
demographic_data <- read_csv("../data/analysis_data/clean_ward_demo_data.csv")

# Identify year columns in budget data
year_cols <- as.character(2022:2031)

# Aggregate total funding per ward number
total_funding_ward <- budget_data |>
  group_by(`Ward Number`)|>
  summarise(
    total_funding = sum(across(all_of(year_cols)), na.rm = TRUE)
  ) |>
  arrange(desc(total_funding))

# Adjust total funding (amounts are in thousands of dollars)
total_funding_ward <- total_funding_ward |>
  mutate(total_funding = total_funding * 1000)

# Pivot demographic data from wide to long format
demographic_long <- demographic_data |>
  pivot_longer(
    cols = -category,
    names_to = "Ward Number",
    values_to = "Value"
  )

# Pivot demographic data back to wide format to get one row per ward
demographic_wide <- demographic_long |>
  pivot_wider(
    names_from = category,
    values_from = Value
  )

# Convert 'Ward Number' to numeric in both datasets to allow merging
total_funding_ward$`Ward Number` <- as.numeric(as.character(total_funding_ward$`Ward Number`))
demographic_wide$`Ward Number` <- as.numeric(as.character(demographic_wide$`Ward Number`))

# Merge budget data with demographic data
merged_data <- total_funding_ward |>
  left_join(demographic_wide, by = "Ward Number")

# Convert 'Total Population' to numeric
merged_data$`Total Population` <- as.numeric(gsub(",", "", merged_data$`Total Population`))

# Create per capita funding variable
merged_data <- merged_data |>
  mutate(per_capita_funding = total_funding / `Total Population`)

# Convert columns from list to double 
merged_data$`Bachelor’s degree or higher` <- as.numeric(merged_data$`Bachelor’s degree or higher`)
merged_data$`Total visible minority population` <- as.numeric(merged_data$`Total visible minority population`)

#Create per capita bachelor's degree variable
merged_data <- merged_data |>
  mutate(`Bachelor’s degree or higher per capita` = `Bachelor’s degree or higher` / `Total Population`)

#Create per capita visible minority population variable
merged_data <- merged_data |>
  mutate(`Total visible minority population per capita` = `Total visible minority population` / `Total Population`)

```

## Data Summary

```{r summary-stats-table, echo=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-demographics-stats
#| tbl-cap: Capital Funding and Socioeconomic Indicators by Ward

# Select relevant columns for the table
merged_table_select <- merged_data |>
  select(
    `Ward Number`,
    per_capita_funding,
    `Median total income of households in 2020 ($)`,
    `Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`,
    `Bachelor’s degree or higher per capita`,
    `Total visible minority population per capita`
  )

# Arrange by the numeric version of per capita funding before formatting
merged_table_select <- merged_table_select |>
  arrange(desc(per_capita_funding))

# Format numeric columns with conditional formatting for LaTeX
merged_table_2 <- merged_table_select |>
  mutate(
    # Format Per Capita Funding as dollars
    per_capita_funding = dollar(per_capita_funding),
    `Median total income of households in 2020 ($)` = dollar(as.numeric(`Median total income of households in 2020 ($)`)),
    `Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)` = paste0(round(as.numeric(`Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`) * 100, 2), "%"),
    `Bachelor’s degree or higher per capita` = paste0(round(as.numeric(`Bachelor’s degree or higher per capita`) * 100, 2), "%"),
    `Total visible minority population per capita` = paste0(round(as.numeric(`Total visible minority population per capita`) * 100, 2), "%")
  )

# Create the table with kable and apply styling
kable(
  merged_table_2, 
  col.names = c(
    "Ward Number", 
    "Per Capita Funding", 
    "Household Median Income", 
    "Low Income Prevalence", 
    "Bachelor’s Degree or Higher", 
    "Visible Minority Population"
  ), 
  table.attr = 'style="width:100%; word-wrap: break-word;"'
) |>
  kable_styling(
    bootstrap_options = c("condensed"), 
    full_width = FALSE,  
    position = "left",  
    font_size = 7,  
  ) |>
  column_spec(1, width = "1cm") |>
  column_spec(2:6, width = "3cm")

```
